Bird beak's elaboration and exploration

### 2022-10-17 Analysis/11-variance_partitioning.Rmd
Update 11-variance_partitioning.Rmd

@@

@@

>>>>

```{r}
library(dispRity)
library(mcmcmcglmmm)
library(dads)
```

@@

Here I'm just doing a one dimensional model of a one dimensional trait (BM) and two clades (`"s"` and `"t"`).

```{r}
## Setting the data
set.seed(0)
data_1 <- drop.fossil.dads(dads(stop.rule = list(max.living = 20), traits = make.traits()))
data_2 <- drop.fossil.dads(dads(stop.rule = list(max.living = 20), traits = make.traits()))

## Renaming the tips and data in data_2
rownames(data_2$data)[match(data_2$tree$tip.label, rownames(data_2$data))] <- gsub("t", "s", data_2$tree$tip.label)
data_2$tree$tip.label <- gsub("t", "s", data_2$tree$tip.label)
data_1$tree$node.label <- data_2$tree$node.label <- NULL

## Combine the trees
ultrametric_tree <- chronos(multi2di(bind.tree(data_1$tree, data_2$tree)))
class(ultrametric_tree) <- "phylo"
tree <- slide.nodes(ultrametric_tree, node = 42, slide = 0.2)

## Combine the data
data <- c(data_1$data[match(data_1$tree$tip.label, rownames(data_1$data)), ], data_2$data[match(data_2$tree$tip.label, rownames(data_2$data)), ])
data <- as.data.frame(matrix(data, ncol = 1, dimnames = list(c(names(data)), c("variable"))))
data <- cbind(data, "var2" = rnorm(40), "clade" = as.factor(c(rep("t", 20), rep("s", 20))), "animal" = rownames(data))
```

We will now run it as the following model:

```
data ~ residuals + random.phylo + random.clade_s + random.clade_t
```

```{r}
## Setting up the model
first_model <- make.mini.chains(data, dimensions = c(1:2), tree = tree, trait.family = "gaussian", residuals = "global", randoms = c("global", "clade"), verbose = TRUE, parameters =  list(nitt = 20000))

## running the first model
first_model_run <- run.mini.chains(first_model, replicates = 1)[[1]]
```

And this is what it looks like graphically

```{r, echo = FALSE}
## Model object
first_model_results <- MCMCglmm.subsets(
      data          = data,
      posteriors    = first_model_run,
      group         = MCMCglmm.levels(first_model_run)[c(1:3)],
      rename.groups = c("clade_s", "clade_t", "phylogeny"))

## Set the clade colours
clade_cols <- c("clade_s" = "blue", "clade_t" = "orange", "phylogeny" = "grey")

## Plotting the model
covar.plot(first_model_results, col = clade_cols, ellipses = mean, legend = TRUE)
```
We can then extract the variance of the model:

```{r}
plot.decompose.var <- function(data, posteriors, group.cols, obs.col = "red", relative = FALSE, ...) {
    ## Get the model variance
    model_variance <- MCMCglmm.variance(posteriors, scale = relative)
    colnames(model_variance) <- c(name.subsets(data), "residuals")
    ## Get the data variance
    if(!relative) {
        obs_var <- sd(unlist(data$matrix[[1]][, c(data$call$dimensions)]))^2
    }
    var_name <- ifelse(relative, "relative variance", "variance")
    if(missing(group.cols)) {
        group_cols <- rep("grey", ncol(model_variance))
    }

    ## Plotting all these with logs
    par(mfrow = c(2,2))

    ## Overall results
    boxplot(model_variance, main = paste0(var_name, " per terms"), xlab = var_name, col = group.cols, horizontal = TRUE, las = 1)
    if(!relative) {
        abline(v = obs_var, col = obs.col, lwd = 2)
    }
    legend("topright", legend = "data variance", col = obs.col, lty = 1, lwd = 2)

    ## Logged results
    boxplot(apply(model_variance, 2, log), main = paste0(var_name, " per terms (logged)"), xlab = paste0(var_name, "(logged)"), col = group.cols, horizontal = TRUE, las = 1)
    if(!relative) {
        abline(v = log(obs_var), col = obs.col, lwd = 2)
    }

    ## Results by sample size
    sample_sizes <- c(size.subsets(data), "residuals" = nrow(data$matrix[[1]]))
    median_variances <- apply(model_variance, 2, median)    
    plot(y = median_variances, x = sample_sizes, ylab = paste0("median ", var_name), xlab = "sample size", pch = 19, col =  group.cols)
    if(!relative) {
        points(y = obs_var, x = sample_sizes[length(sample_sizes)], col = "red", pch = 19)
    }
    model <- lm(median_variances ~ sample_sizes)
    abline(model$coefficients)
    fit <- paste0("lm fit (adj.R^2) = ", round(summary(model)$adj.r.squared, 3))
    legend("topright", legend = fit, lty = 1)

    ## Results log logged
    sample_sizes <- c(size.subsets(data), "residuals" = nrow(data$matrix[[1]]))
    median_variances <- apply(model_variance, 2, median)    
    plot(y = log(median_variances), x = log(sample_sizes), ylab = paste0("median ", var_name, " (log)"), xlab = "sample size (log)", pch = 19, col =  group.cols)
    if(!relative) {
        points(y = log(obs_var), x = log(sample_sizes[length(sample_sizes)]), col = "red", pch = 19)
    }
    model <- lm(log(median_variances) ~ log(sample_sizes))
    abline(model$coefficients)
    fit <- paste0("lm fit (adj.R^2) = ", round(summary(model)$adj.r.squared, 3))
    legend("topright", legend = fit, lty = 1)
}

plot.decompose.var(first_model_results, first_model_run, group.cols = c(clade_cols, "residuals" = "darkgrey"), relative = FALSE)
```

@@

For this model we can use the charadriiformes example data from `dispRity` where the groups are slightly different sizes (but equal-ish) but based on biological data.

```{r}
data(charadriiformes)
## Same but selecting only the three first random terms
chara_data <- MCMCglmm.subsets(data       = charadriiformes$data,
                               posteriors = charadriiformes$posteriors,
                               tree       = charadriiformes$tree,
                               group      = MCMCglmm.levels(charadriiformes$posteriors)[1:4],
                               rename.groups = c("gulls", "plovers", "sandpipers", "phylogeny"))
```

This is what it looks like:

```{r}
## Set up the clade colours
clade_cols <- c("gulls" = "blue", "plovers" = "orange", "sandpipers" = "darkgreen", "phylogeny" = "grey")
## Plot the model
covar.plot(chara_data, col = clade_cols, ellipses = mean, legend = TRUE, lwd = 2, points.cex = 0.5)
```

And this is what the variances look like:

```{r}
plot.decompose.var(chara_data, charadriiformes$posteriors, group.cols = c(clade_cols, "residuals" = "darkgrey"))
```
